---
title: "Introduction to R"
author: "Clay Ford"
date: "Fall 2016"
output: beamer_presentation
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 50)
```

## Today's workshop

Teach you how to:

- read in and manipulate tabular data
- do some basic statistics
- do some graphing

I'm assuming you're new to R.  
\ 

The goal: get you comfortable enough to help yourself go further with R.


## Introducing R

- R is an environment for statistical computing and graphics that has a programming language
- Freely available and maintained by volunteers
- R is extensible; can be expanded by installing "packages" or by creating your own "functions"
- Frequently used with RStudio, a free IDE (Interactive Development Environment) for R


## How to get R and RStudio

**R**: http://cran.r-project.org/ (or better yet, Google “Download R”)  
\ 

**RStudio**: http://www.rstudio.com/   
\ 

Both are available for Windows, Mac, and Linux and free to install (no catches).  
\ 

**If you install R and RStudio, then you only need to run RStudio.**



## R basics - functions

R uses *functions* to perform operations.  
\ 

Functions take *arguments* to specify how operations are performed.   
\ 

Example: `read.csv(file="scores.csv")`   
\ 

`read.csv` is a function to import a CSV file; `file` is an argument.   
\ 

R has **many** functions. Fortunately about 20% of them do about 80% of the work.


## R basics - running functions

You can run functions two ways:

1. In the console at the command line; type the function and hit Enter
2. In an R script; type the function and (with cursor on same line as code) hit Ctrl + Enter (Command + Enter on Mac)

An R script is a text file that contains all your R code. R scripts allow you to save, edit, reproduce and share your code.   
\ 

Today's workshop will be centered around an R script I wrote in advance.



## R basics - assignment

We often need to save a function's result or output. For this we use the assignment operator: ` <- `  
\ 

For example, when you import a CSV file you need to give it a name:  
\ 

`scoreData <- read.csv(file="scores.csv")`   
\ 

Now we can use `scoreData` as an argument to other functions. For example: `summary(scoreData)` which computes summaries for each column in the data.   
\ 

**Note**: Use `Alt + -` (Win) or `Option + -` (Mac) in RStudio to quickly insert ` <- `. You can also use `=` for assignment.

## R basics - objects

In R we create and manipulate *objects*. Examples of objects are vectors, matrices, data frames and lists. You can create as many objects as memory allows.   
\ 

From the previous slide:   
\ 

`scoreData <- read.csv(file="scores.csv")`   
\ 

This creates a data frame object called `scoreData`.

## R objects - vectors

One-dimensional objects consisting of the same things (numbers or characters).

```{r, collapse=TRUE}
# create vector of character strings
gender <- c("Male","Female","Male") 
gender

# create vector of numbers
x <- seq(0, 40, by=2) 
x
```

## R objects - matrices

Two-dimensional objects consisting of the same things (numbers or characters).

```{r}
M <- matrix(1:4,ncol=2)
M
```

Arrays are another type of object. Like matrix, but multi-dimensional.

## R objects - data frames

Two-dimensional objects consisting of different things (numbers or characters), but the same number of things.

```{r}
df <- read.csv("C:/data/grades.csv")
df
```


## R objects - lists

An object consisting of different things (numbers or characters), and with different numbers of those things.

```{r}
x <- 1:100
y <- list(Length = length(x), Summary = summary(x),
          Comment = "This is x.")
y
```

Statistical output is often contained in lists. 

## R basics - packages

All functions belong to *packages*. R comes with about 30 packages, but as of August 2016, there are *over 8700* user-contributed packages!   
\ 

To use functions in a package, the package must be installed and loaded.    
\ 

You only _install_ a package once. You _load_ a package whenever you want to use its functions. You also sometimes need to _update_ packages when new versions are released.      
\ 

Examples of packages that come with R:

- The `stats` package contains statistical functions
- The `foreign` package contains functions for reading data from other statistical programs
- The `graphics` package contains functions for creating graphics


## R in a Nutshell

**Use functions in various packages to create and manipulate objects**.  
\ 

So we need to learn...

- some useful functions
- how to install and load packages
- about different R objects
- how to use RStudio
- how to get help

Let's open up RStudio and get started!


## Keep Going!

- Join the UVa R Users Group: http://www.meetup.com/UVa-R-Users-Group/

- Sign up for the R-Bloggers daily email: http://www.r-bloggers.com/

- Lander, J., *R for Everyone*, Addison-Wesley, 2014.

- Teetor, P., *R Cookbook*, O'Reilly, 2011.

- Coursera's *Computing for Data Analysis*: https://www.coursera.org/course/compdata


## Check out these web sites

- Contributed Documentation: http://cran.r-project.org/other-docs.html

- Quick-R: http://www.statmethods.net/

- Cookbook for R: http://www.cookbook-r.com/

- R-statistics blog: http://www.r-statistics.com/

- The R Journal: http://journal.r-project.org/

- UCLA R Starter Kit: http://www.ats.ucla.edu/stat/r/sk/


## StatLab

Thanks for coming today!  
\ 

For help and advice with your data analysis, contact the StatLab to set up an appointment: statlab@virginia.edu  
\ 

Sign up for more workshops or see past workshops:
http://data.library.virginia.edu/training/   
\ 

Register for the Research Data Services newsletter to stay up-to-date on StatLab events and resources: http://data.library.virginia.edu/newsletters/
